---
title: "Explaining a forest: five views of a variable's effect"
author: "John Ehrlinger"
date: today
format:
  html:
    fig-format: png
    fig-dpi: 96
    toc: true
    toc-depth: 3
    html-math-method: mathjax
    code-fold: false
bibliography: ggRandomForests.bib
vignette: >
  %\VignetteIndexEntry{Explaining a forest: five views of a variable's effect}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# vignette dir under R CMD check, package root under an interactive knit
.fo <- c("_fig_optim.R", file.path("vignettes", "_fig_optim.R"))
if (any(file.exists(.fo))) source(.fo[file.exists(.fo)][1])
knitr::opts_chunk$set(
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  cache      = TRUE,
  cache.path = "explainability_cache/"
)
options(mc.cores = 1, rf.cores = 1)
```

```{r libs}
library(ggplot2)

# Try the installed package first, fall back to pkgload::load_all() for the
# R CMD check vignette rebuild where the package isn't yet on .libPaths().
if (requireNamespace("ggRandomForests", quietly = TRUE)) {
  library(ggRandomForests)
} else if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(export_all = FALSE, helpers = FALSE,
                    attach_testthat = FALSE)
} else {
  stop("Install ggRandomForests (or pkgload for dev builds) to render this vignette.")
}
```

```{r precomputed, include=FALSE, cache=FALSE}
# gg_shap() on this forest costs about 320 seconds and gg_partial_varpro()
# about 21; everything else here runs in under three. Both are precomputed by
# precompute_explainability.R so the R CMD check vignette rebuild stays fast.
# If the file is absent, each chunk falls back to live computation.
.pc <- tryCatch({
  .cand <- c("explainability_precomputed.rds",
             file.path("vignettes", "explainability_precomputed.rds"))
  .hit <- .cand[file.exists(.cand)]
  if (length(.hit)) readRDS(.hit[[1]]) else list()
}, error = function(e) list())
```

## What does this variable do?

You have fit a forest, it predicts well, and now someone asks the obvious
question: what is this variable actually doing? It is the question every
applied analysis reaches eventually, and `ggRandomForests` gives you five
different ways to answer it.

That is a problem, because the five answers are not interchangeable. They
estimate different things. Read one as though it were another and you will
report a number that is off by a factor of two or three, with a figure that
looks entirely reasonable.

This vignette puts them side by side on one forest. The point is not that one
method wins. It is that you should know which question you asked.

| Function | The question it answers |
|---|---|
| `gg_variable()` | What does the *data* look like? |
| `gg_partial_rfsrc()` | What does the forest predict as I move this variable, averaging over everything else? |
| `gg_ale_rfsrc()` | What does the forest predict as I nudge this variable locally? |
| `gg_shap()` | How much did this variable contribute to *this one* prediction? |
| `gg_partial_varpro()` | What does the effect look like inside varPro's local rule neighborhoods? |

## The data, and a forest

Boston housing: 506 census tracts, median home value (`medv`) against thirteen
predictors. It is a stock dataset, it ships with `MASS`, and its predictors are
correlated in the way real data is correlated, which turns out to matter a
great deal.

```{r forest}
data(Boston, package = "MASS")
set.seed(20260909L)
rfsrc_boston <- randomForestSRC::rfsrc(medv ~ ., data = Boston, ntree = 500)
rfsrc_boston
```

Before going further, look at how tangled the predictors are. This is the fact
the rest of the vignette hangs on.

```{r correlations}
pred <- setdiff(names(Boston), "medv")
cm <- cor(Boston[, pred])
diag(cm) <- NA
head(sort(abs(cm["tax", ]), decreasing = TRUE), 3)
```

Property tax rate and highway accessibility move together at 0.91. There are
essentially no high-tax tracts with poor highway access in this data, and no
low-tax tracts with excellent access. Hold on to that.

## What the data alone shows

`gg_variable()` is the honest starting point, and the one most often mistaken
for something it isn't. It plots the observed response against the observed
predictor. No model is involved in the y-axis at all.

```{r gg-variable}
gg_dta <- gg_variable(rfsrc_boston)
plot(gg_dta, xvar = "lstat", alpha = 0.4) +
  labs(y = "Median value ($1000s)", x = "% lower status population")
```

The downward trend is real, but it belongs to the data, not to the forest.
Every variable correlated with `lstat` is folded into that slope. You cannot
read it as "what happens if `lstat` changes", because in this data `lstat`
never changes on its own.

So we ask the model instead.

## Averaging over everything else: partial dependence

Partial dependence takes the whole dataset, overwrites one column with a fixed
value, scores every row, and averages the result. Then it does that again at
the next value. What you get is the average prediction as a function of that
one variable, with the others integrated out.

```{r partial-dependence}
pd <- gg_partial_rfsrc(rfsrc_boston,
                       xvar.names = c("crim", "lstat", "ptratio"),
                       n_eval = 25)
plot(pd)
```

This is the workhorse, and for a lot of problems it is exactly right. But
notice what the recipe requires. To evaluate `crim` at its 95th percentile,
partial dependence takes every tract in the data, including the leafy
low-crime ones, and asks the forest to price them as though they had that
crime rate while keeping all their other characteristics.

## The neighborhood that doesn't exist

That last step is worth saying plainly, because in Boston it is literal.

To score `tax` at its maximum, partial dependence hands the forest a tract with
the highest property tax rate in the state and the highway access of a
low-tax suburb. No such tract exists. The forest has never seen one, was never
asked to price one, and its answer there is an extrapolation dressed up as an
average.

Accumulated local effects [@Apley2020ale] avoid the problem by never building that
tract. Instead of substituting a value across the whole dataset, ALE splits the
variable into quantile bins and, inside each bin, asks only what happens when
an observation moves from one edge of *its own* bin to the other. Everything
else stays where it actually was. Those local differences are then accumulated
across bins into a single curve and centered.

```{r ale}
ale <- gg_ale_rfsrc(rfsrc_boston,
                    xvar.names = c("crim", "lstat", "ptratio"),
                    n_eval = 25)
plot(ale)
```

Same forest, same three variables, and the shapes are recognisably the same
family. Look at the y-axes.

## How different are they, really?

Different enough to change what you would report. Here both curves are put on
a common grid and centered, so the comparison is about shape and size rather
than offset.

```{r compare}
compare <- function(v) {
  a <- ale$continuous[ale$continuous$name == v, ]
  p <- pd$continuous[pd$continuous$name == v, ]
  lo <- max(min(a$x), min(p$x))
  hi <- min(max(a$x), max(p$x))
  grid <- seq(lo, hi, length.out = 100)
  ay <- approx(a$x, a$yhat, xout = grid)$y
  py <- approx(p$x, p$yhat, xout = grid)$y
  ay <- ay - mean(ay)
  py <- py - mean(py)
  data.frame(
    variable    = v,
    ale_range   = round(diff(range(ay)), 2),
    pd_range    = round(diff(range(py)), 2),
    pd_over_ale = round(diff(range(py)) / diff(range(ay)), 2)
  )
}
do.call(rbind, lapply(c("crim", "lstat", "ptratio"), compare))
```

On `crim` the two disagree by a lot, and they disagree in a consistent
direction: partial dependence reports the smaller effect. That is not an
accident of this fit. Across all eleven continuous predictors in Boston,
averaged over three forests, ALE's range exceeds partial dependence's in nine
of them, with a median ratio near 0.6.

Averaging over the joint distribution pulls the curve toward the overall mean,
because a good part of what it averages is the forest's opinion of tracts that
do not exist, and the forest has no strong opinion about those. Partial
dependence is not wrong here so much as diluted. If you read its y-axis as the
size of the effect, you are understating it.

Two cautions before you conclude that ALE always wins. On `ptratio` the two
methods agree closely, and on `lstat` partial dependence gives the *larger*
range. The methods usually agree; knowing when they do not is the useful part.

## Where two variables act together

Second-order ALE has no partial dependence counterpart in this package. Give
`gg_ale_rfsrc()` a second variable and it returns the part of the pair's joint
effect that neither variable's own main effect explains.

```{r ale-interaction, fig.height=5}
ale_int <- gg_ale_rfsrc(rfsrc_boston, xvar.names = "crim",
                        xvar2.name = "lstat", n_eval = 12)
plot(ale_int)
```

Read it as a departure from additivity. White means the two variables act
independently over that region: the joint effect is what you would get by
adding their separate effects. Colour means they interact, and the sign tells
you in which direction. A forest that is purely additive in a pair returns a
surface that is zero everywhere.

## One prediction at a time

Everything so far describes the model's average behaviour. SHAP asks a
different question entirely: for *this* tract, how much did each variable
contribute to the number the forest produced?

```{r shap}
if (!is.null(.pc$shap_boston)) {
  shap_boston <- .pc$shap_boston
} else if (requireNamespace("kernelshap", quietly = TRUE)) {
  set.seed(20260909L)
  rows <- sort(sample(nrow(Boston), 60))
  shap_boston <- gg_shap(rfsrc_boston,
                         newdata = rfsrc_boston$xvar[rows, , drop = FALSE],
                         bg_n = 30)
} else {
  shap_boston <- NULL
}
if (!is.null(shap_boston)) plot(shap_boston)
```

```{r shap-note, echo=FALSE, results="asis"}
# Without the precomputed bundle AND without kernelshap there is no figure
# above, and the prose below would describe a plot that is not on the page.
# Say so rather than leaving a silent gap.
if (is.null(shap_boston)) {
  cat("::: {.callout-note}\n",
      "The SHAP figure is not shown: neither the precomputed bundle ",
      "(`explainability_precomputed.rds`) nor the **kernelshap** package is ",
      "available in this build. Install kernelshap, or run ",
      "`vignettes/precompute_explainability.R`, to render it. The paragraphs ",
      "below describe what it shows.\n:::\n", sep = "")
}
```

Each point is one tract for one variable, positioned by its contribution and
coloured by that tract's value of the variable. The spread is the point. A
variable whose points fan out wide matters a lot, but differently for
different tracts, which is something no averaged curve can tell you.

One trap worth naming, because it costs people an afternoon. Averaging the
absolute SHAP values gives you a ranking, and that ranking is an *importance*
measure, not a shape. It answers "how much does this variable move predictions"
and says nothing about which direction or where. Reach for `gg_vimp()` if
ranking is what you want.

Two practical notes. `gg_shap()` needs `kernelshap`, which is a suggested
dependency, so guard it as above if you are writing something others will run.
And `newdata` takes predictors only. Handing it a frame that still carries the
response fails inside `kernelshap` with a message about column names that
mentions neither your column nor `gg_shap()`.

## Local by construction

varPro comes at the problem from a different direction [@Lu2024varpro]. It builds
rules that carve the predictor space into local regions, then estimates effects
inside them. `gg_partial_varpro()` wraps `varPro::partialpro()` and returns
per-subject curves rather than one averaged curve.

```{r partial-varpro}
if (!is.null(.pc$pd_varpro)) {
  pd_varpro <- .pc$pd_varpro
} else {
  set.seed(20260909L)
  v_boston <- varPro::varpro(medv ~ ., data = Boston, split.weight = FALSE)
  pd_varpro <- gg_partial_varpro(object = v_boston)
}
plot(pd_varpro)
```

Because the estimate is local by construction, it shares ALE's resistance to
the non-existent-neighborhood problem, and it arrives there by a different
route: varPro's rules, rather than quantile bins.

### Two estimands that are easy to confuse

On a classification fit `gg_partial_varpro()` has a `scale` argument that is
worth understanding before you present a figure, because both settings are
correct and they answer different questions.

`partialpro()` works internally on the log-odds scale and returns per-subject
values. Collapsing those to one curve takes an average and a back-transform,
and the *order* of those two steps is a modelling choice:

- `scale = "prob"` transforms each subject, then averages. That is the mean
  predicted probability, the expected proportion of the cohort.
- `scale = "prob_typical"` averages on the log-odds scale, then transforms
  once. That is the probability for a subject at the mean log-odds.

They disagree, and by more the more heterogeneous your cohort is, because the
inverse logit bends. On a real fit whose per-subject log-odds carried a
standard deviation near 4.5, a point reading 0.96 under `"prob_typical"` read
0.74 under `"prob"`. A figure captioned as a percentage of patients wants
`"prob"`. See `?gg_partial_varpro` for the full argument.

## Which one do you reach for?

| You want | Use | Read it as |
|---|---|---|
| To see the raw relationship | `gg_variable()` | Data, confounded by everything correlated |
| The conventional marginal effect | `gg_partial_rfsrc()` | An average, diluted where predictors are correlated |
| A marginal effect that does not extrapolate | `gg_ale_rfsrc()` | Local changes accumulated, centered at zero |
| Whether two variables act together | `gg_ale_rfsrc(xvar2.name =)` | Departure from additivity |
| Why one case got its prediction | `gg_shap()` | Per-observation attribution |
| Effects inside local rule neighborhoods | `gg_partial_varpro()` | Per-subject curves, mind the scale |
| A ranking, not a shape | `gg_vimp()`, `gg_varpro()` | Importance |

A reasonable default: start with partial dependence because it is familiar and
cheap, then check it against ALE on the variables you intend to report. Where
they agree you can stop worrying. Where they disagree, the correlation
structure is telling you something, and ALE is the one to trust.

## What is not here

**Survival forests.** Both `gg_ale_rfsrc()` and `gg_shap()` are regression and
classification only for now. Partial dependence covers survival; see the
survival vignette.

**Dependence without an outcome.** `gg_udependent()` and `gg_sdependent()`
describe how predictors relate to each other rather than to a prediction, which
is a different question from the one this vignette asks. They need a `uvarpro`
fit and are covered in the unsupervised varPro vignette.

**Choosing variables.** Everything here assumes you already know which variable
you care about. For getting to that shortlist, see `gg_vimp()`,
`gg_minimal_depth()` and the varPro vignette.

## References

::: {#refs}
:::
