---
title: "Exploring Random Forests with ggRandomForests"
author: "John Ehrlinger"
date: today
format: 
  html:
    toc: true
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 80
vignette: >
  %\VignetteIndexEntry{Vignette's Title}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **ggRandomForests** package extracts tidy data objects from either
`randomForestSRC` or `randomForest` fits and feeds them into familiar
`ggplot2` workflows. This vignette highlights the most common objects—
`gg_error`, `gg_variable`, and `gg_vimp`—along with a small helper for
building balanced conditioning intervals.

## Error trajectories with `gg_error()`

```{r error-demo}
library(ggRandomForests)
library(randomForest)
set.seed(42)
rf_iris <- randomForest(Species ~ ., data = iris, ntree = 200, keep.forest = TRUE)
err_df <- gg_error(rf_iris, training = TRUE)
head(err_df)
```

The `gg_error()` object stores the cumulative OOB error rate for each
outcome column plus the `ntree` counter. When `training = TRUE`, the
function reconstructs the original model frame and appends the in-bag
error trajectory (`train`). Plotting overlays both curves by default:

```{r error-plot, fig.height=4}
plot(err_df)
```

## Marginal dependence via `gg_variable()`

```{r variable-demo}
set.seed(99)
boston <- MASS::Boston
rf_boston <- randomForest(medv ~ ., data = boston, ntree = 150)
var_df <- gg_variable(rf_boston)
str(var_df[, c("lstat", "yhat")])
```

Because the original training data are recovered from the model call,
`gg_variable()` works even when the forest was trained within helper
functions or against a `subset()` expression. The output keeps the raw
predictors plus either a continuous `yhat` column (regression) or per-class
probabilities (`yhat.<class>` for classification). Plotting a single
variable is straightforward:

```{r variable-plot, fig.height=4}
plot(var_df, xvar = "lstat")
```

Survival forests can request multiple horizons using the `time` argument;
non-OOB predictions are available by setting `oob = FALSE`.

## Variable importance with `gg_vimp()`

```{r vimp-demo}
vimp_df <- gg_vimp(rf_boston)
head(vimp_df)
plot(vimp_df)
```

If a `randomForest` object lacks stored importance scores, `gg_vimp()`
tries to compute them on the fly. When the forest truly cannot provide the
information (for example when `importance = FALSE` and the predictors are
no longer accessible), the function emits a warning and returns `NA`
placeholders so plots still render.

## Balanced conditioning cuts with `quantile_pts()`

```{r quantile-demo}
rm_breaks <- quantile_pts(boston$rm, groups = 6, intervals = TRUE)
rm_groups <- cut(boston$rm, breaks = rm_breaks)
table(rm_groups)
```

The helper wraps `stats::quantile()` to produce evenly populated strata
that drop directly into `cut()` when building coplots or facet labels.

## Next steps

* Inspect the full API reference at <https://ehrlinger.github.io/ggRandomForests/>.
* Use `?gg_error`, `?gg_variable`, `?gg_vimp`, and `?quantile_pts` for
  additional arguments and examples.
* Pair these data objects with your own `ggplot2` themes to align with your
  preferred publication style.
