---
title: "Random Forest Regression with ggRandomForests"
author: "John Ehrlinger"
date: today
format:
  html:
    fig-format: png
    fig-dpi: 96
    toc: true
    toc-depth: 3
    html-math-method: mathjax
    code-fold: false
bibliography: ggRandomForests.bib
vignette: >
  %\VignetteIndexEntry{Random Forest Regression with ggRandomForests}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

::: {.callout-warning}
## Work in progress
This vignette is under active development. Code examples and narrative may
change before the next release.
:::

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
options(mc.cores = 1, rf.cores = 1)
```

# Introduction

Random forests [@Breiman:2001] are a non-parametric ensemble method that
requires no distributional or functional assumptions on how covariates relate to
the response. The method builds a large collection of de-correlated decision
trees via bootstrap aggregation (bagging) and random feature selection, then
averages their predictions to smooth out the noise any single tree carries. The
**randomForestSRC** package [@Ishwaran:RFSRC:2014] provides a unified
implementation for survival, regression, and classification forests.

**ggRandomForests** extracts tidy data objects from `rfsrc` fits and renders
them with **ggplot2** [@Wickham:2009], making it straightforward to explore how a
forest is constructed, which variables matter, and how the response depends on
individual predictors.

This vignette demonstrates a complete random forest regression workflow on the
Boston Housing data set [@Harrison:1978; @Belsley:1980]:

1. **Data exploration** --- EDA scatter panels, variable descriptions
2. **Growing the forest** --- fitting an RF, checking OOB error convergence
3. **Variable selection** --- VIMP and minimal depth via `max.subtree()`
4. **Dependence plots** --- variable dependence and partial dependence via
   `gg_variable()` and `gg_partial_rfsrc()`
5. **Variable interactions** --- conditioning plots and partial dependence
   surfaces

```{r packages}
library(ggplot2)
library(dplyr)
library(tidyr)
library(randomForestSRC)

if (requireNamespace("ggRandomForests", quietly = TRUE)) {
  library(ggRandomForests)
} else if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, attach_testthat = FALSE)
} else {
  stop("Install ggRandomForests (or pkgload for dev builds) to render this vignette.")
}

theme_set(theme_bw())
```

# Data: Boston Housing Values

The Boston Housing data [@Harrison:1978; @Belsley:1980] is a standard benchmark
for regression. It contains data for 506 census tracts of Boston from the 1970
census. The objective is to predict the median value of owner-occupied homes
(`medv`, in \$1000s) from 13 predictors covering crime, zoning, industry,
environmental quality, and housing characteristics. We use the copy from the
**MASS** package [@mass:2002].

```{r data-load}
data(Boston, package = "MASS")
Boston$chas <- as.logical(Boston$chas) # nolint: object_name_linter
```

```{r variable-labels}
st_labs <- c(
  crim    = "Crime rate by town",
  zn      = "Residential land zoned > 25k sq ft (%)",
  indus   = "Non-retail business acres (%)",
  chas    = "Borders Charles River",
  nox     = "Nitrogen oxides (10 ppm)",
  rm      = "Rooms per dwelling",
  age     = "Units built before 1940 (%)",
  dis     = "Distance to employment centers",
  rad     = "Highway accessibility index",
  tax     = "Property tax rate per $10,000",
  ptratio = "Pupil-teacher ratio",
  black   = "Proportion of Black residents",
  lstat   = "Lower status population (%)",
  medv    = "Median home value ($1000s)"
)
```

## Exploratory data analysis

We plot each predictor against the response (`medv`), coloring by the sole
categorical variable (`chas`, whether the tract borders the Charles River). A
loess smooth highlights the marginal trend.

```{r eda, fig.cap="EDA: each predictor vs. median home value, colored by Charles River indicator."}
dta <- Boston |>
  pivot_longer(c(-medv, -chas), names_to = "variable", values_to = "value")

ggplot(dta, aes(x = medv, y = value, color = chas)) +
  geom_point(alpha = 0.4) +
  geom_smooth(aes(x = medv, y = value), color = "grey30",
              inherit.aes = FALSE, se = FALSE) +
  labs(y = "", x = st_labs["medv"]) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~variable, scales = "free_y", ncol = 3)
```

Even from this simple view, two relationships stand out: `medv` against
`lstat` (lower status %) and `medv` against `rm` (rooms per dwelling). Keep
those two in mind --- we expect the random forest to rank them as the most
important predictors, and the rest of the vignette comes back to check.

# Growing a Random Forest

We grow a regression forest using all 13 predictors. The `rfsrc()` function
detects the regression family from the continuous response.

```{r rfsrc-fit}
rfsrc_Boston <- rfsrc(medv ~ ., data = Boston, # nolint: object_name_linter
                      ntree = 200, importance = TRUE, err.block = 5)
rfsrc_Boston
```

The forest grew `r rfsrc_Boston$ntree` trees, splitting on
`r rfsrc_Boston$mtry` randomly selected candidate variables at each node, and
stopping at a minimum terminal node size of `r rfsrc_Boston$nodesize`.

## OOB error convergence

```{r error-plot, fig.height=4, fig.cap="OOB mean squared error vs. number of trees."}
gg_e <- gg_error(rfsrc_Boston)
gg_e <- gg_e |> filter(!is.na(error))
class(gg_e) <- c("gg_error", class(gg_e))
plot(gg_e)
```

The error stabilizes well before 500 trees, indicating the forest is large
enough for reliable predictions.

## OOB predictions

```{r rfsrc-predicted, fig.height=4, fig.cap="OOB predicted median home values. Points are jittered; boxplot shows the distribution."}
plot(gg_rfsrc(rfsrc_Boston), alpha = 0.5) +
  coord_cartesian(ylim = c(5, 49))
```

Each point is a single tract's OOB prediction. The distribution is a sanity
check --- we are more interested in the *why* behind these predictions.

# Variable Selection

## Variable importance (VIMP)

VIMP, computed by [randomForestSRC](https://www.randomforestsrc.org), measures
the increase in OOB prediction error when a variable's values are randomly
permuted across the out-of-bag observations [@Breiman:2001]. Permutation severs
the variable's link to the response on purpose: if breaking that link hurts
accuracy, the variable is carrying real signal. Large positive values mean the
variable is essential; negative values suggest it is no more informative than
noise.

```{r vimp-plot, fig.cap="VIMP ranking. Longer blue bars indicate more important variables."}
plot(gg_vimp(rfsrc_Boston), lbls = st_labs)
```

`lstat` and `rm` dominate, with a clear gap to the remaining predictors. All
VIMP values are positive, indicating every predictor contributes at least
marginally.

The permutation approach contrasts with varPro release-rule importance
[@Lu2024varpro], available through `gg_varpro()`. Rather than perturbing data
synthetically, varPro compares local estimators on the observed data directly:
no permutation, no manufactured feature values. Because the two methods measure
fundamentally different things, a variable can rank high under one and low under
the other. When they agree, the evidence is strong; when they disagree, that
disagreement itself is worth investigating, pointing either to a variable whose
effect is highly non-linear or to one that matters only in combination with
others.

## Minimal depth

Minimal depth [@Ishwaran:2010] ranks variables by how close to the root node
they first split, on average. Variables that partition large portions of the
population early are considered most important.

```{r varsel}
md_Boston <- max.subtree(rfsrc_Boston) # nolint: object_name_linter
```

The threshold is `r round(md_Boston$threshold, 2)`, selecting
`r length(md_Boston$topvars)` variables:
`r paste(md_Boston$topvars, collapse = ", ")`.

Both VIMP and minimal depth agree on the dominance of `lstat` and `rm`. We use
the minimal depth top variables for the remainder of the analysis.

```{r select-vars}
xvar <- md_Boston$topvars
```

# SHAP Analysis

VIMP and varPro both rank *how much* a variable matters, averaged over the
whole forest. They cannot tell you how much a variable mattered for one
specific tract's prediction --- that requires a different kind of
accounting. SHAP (SHapley Additive exPlanations) borrows an idea from
cooperative game theory: treat the 13 predictors as players splitting a
payout, and ask how much each one contributed to this tract's predicted
`medv`, averaged fairly over every order the players could have joined the
game. The result is a signed contribution per predictor per tract, and those
contributions add up exactly: the forest's overall average prediction (the
*baseline*) plus the sum of one tract's contributions equals that tract's
actual predicted value. No other importance measure in this vignette makes
that promise.

`gg_shap()` computes these contributions by calling
[kernelshap](https://cran.r-project.org/package=kernelshap), which needs one
prediction per predictor per background draw, per tract explained. Scoring
all 506 tracts is expensive, so we explain a random sample of 40 instead,
plenty to see the patterns at a fraction of the cost.

```{r shap-fit}
set.seed(42)
shap_sample <- Boston[sample(nrow(Boston), 40), setdiff(colnames(Boston), "medv")]
gg_shp <- gg_shap(rfsrc_Boston, newdata = shap_sample, bg_n = 50)
```

## SHAP importance

Averaging the absolute contribution of each variable across the 40 tracts
gives a ranking directly comparable to VIMP.

```{r shap-importance-plot, fig.cap="Mean absolute SHAP value per predictor."}
plot(gg_shp, type = "importance")
```

`lstat` and `rm` come out on top again, same as VIMP and minimal depth ---
three different mechanisms, one answer. That agreement is reassuring, but
it's the next two plots where SHAP earns its keep.

## SHAP beeswarm

The beeswarm plot is one dot per tract per variable: its horizontal position
is the SHAP contribution (negative pulls the prediction down, positive pushes
it up), and its color is that tract's value for the variable, scaled low to
high within each row so the pattern doesn't get washed out by variables on
very different scales.

```{r shap-beeswarm-plot, fig.cap="SHAP beeswarm: every dot is one tract's contribution for one predictor."}
plot(gg_shp, type = "beeswarm")
```

`lstat` shows a clean color gradient: the yellow (high lower-status
population) dots sit on the negative side, the purple (low) dots on the
positive side. `rm` runs the other way. Both track what the EDA scatterplot
already hinted at, but now stated as a contribution to an individual
prediction rather than a marginal trend.

## SHAP dependence

Plotting one variable's SHAP contribution against its own value is the
closest SHAP gets to a partial dependence plot, tract by tract instead of
averaged.

```{r shap-dependence-plot, fig.cap="SHAP dependence for lstat."}
plot(gg_shp, type = "dependence", xvar = "lstat")
```

The downward slope matches the sign of the correlation directly: tracts with
more lower-status population get pulled below the baseline prediction, tracts
with less get pushed above it. The next section asks the same question a
second way, partial dependence averaged over the whole forest instead of
tract by tract, and the two views should agree on the shape even though
they're built from entirely different mechanics.

# Variable Dependence

## Variable dependence plots

Variable dependence shows each tract's OOB predicted `medv` plotted against a
predictor, with a loess smooth indicating the trend.

```{r vardep-panel, fig.cap="Variable dependence for top predictors (minimal depth rank order)."}
gg_v <- gg_variable(rfsrc_Boston)

plot(gg_v, xvar = xvar, panel = TRUE, alpha = 0.5) +
  labs(y = st_labs["medv"], x = "")
```

The panels confirm what EDA suggested: `medv` decreases sharply with `lstat` and
increases with `rm`, both in strongly non-linear ways. The remaining variables
show weaker but still discernible trends.

```{r vardep-chas, fig.height=4, fig.cap="Variable dependence for Charles River (categorical)."}
plot(gg_v, xvar = "chas", alpha = 0.4) +
  labs(y = st_labs["medv"])
```

Most tracts do not border the Charles River, and the predicted value
distributions largely overlap --- consistent with `chas` ranking last in both
VIMP and minimal depth.

## Partial dependence

Partial dependence integrates out the effects of all other covariates, giving a
risk-adjusted view of each predictor's marginal effect [@Friedman:2000]:

$$
\tilde{f}(x) = \frac{1}{n} \sum_{i=1}^n \hat{f}(x, x_{i,o})
$$

We use `gg_partial_rfsrc()`, which calls `randomForestSRC::partial.rfsrc()`
directly and returns a `gg_partial_rfsrc` object. The quickest path to a
figure is `plot(pd)`, which sorts out continuous and categorical variables
for you. When you want to control the layout yourself, the underlying data
is in `pd$continuous`.

```{r partial-dep, fig.cap="Partial dependence for top predictors."}
pd <- gg_partial_rfsrc(rfsrc_Boston, xvar.names = xvar)

# Quick S3 plot — works out of the box for the standard regression case
plot(pd)
```

For a publication-ready layout with custom axis labels, access the underlying
data frame directly:

```{r partial-dep-custom, fig.cap="Partial dependence (custom styling)."}
ggplot(pd$continuous, aes(x = x, y = yhat)) +
  geom_line(color = "steelblue", linewidth = 1) +
  facet_wrap(~name, scales = "free_x") +
  labs(y = st_labs["medv"], x = "") +
  theme_bw()
```

`lstat` shows a strongly concave relationship, while `rm` stays flat below
about 6 rooms and then climbs sharply. Shapes like these are awkward to
capture with a simple parametric transform, since you would have to guess
the form in advance, but the random forest picks them up on its own.

# Variable Interactions and Conditioning Plots

## Conditioning on a categorical variable

The simplest coplot conditions on a categorical variable. Here we examine
`medv` vs. `lstat`, split by Charles River status:

```{r coplot-chas, fig.height=4, fig.cap="Variable dependence on lstat, conditional on Charles River."}
gg_v$chas_label <- ifelse(gg_v$chas, "Borders Charles River",
                          "Does not border")

plot(gg_v, xvar = "lstat", alpha = 0.5) +
  labs(y = st_labs["medv"], x = st_labs["lstat"]) +
  theme(legend.position = "none") +
  facet_wrap(~chas_label)
```

The decreasing trend holds in both groups, with slightly higher values along the
Charles River at every `lstat` level.

## Conditioning on a continuous variable

To investigate the `lstat`--`rm` interaction, we bin `rm` into six quantile
groups using `quantile_pts()` and facet:

```{r coplot-rm, fig.cap="Predicted medv vs. lstat, conditional on rooms-per-dwelling groups."}
rm_pts <- quantile_pts(rfsrc_Boston$xvar$rm, groups = 6, intervals = TRUE)
gg_v$rm_grp <- cut(rfsrc_Boston$xvar$rm, breaks = rm_pts)
levels(gg_v$rm_grp) <- paste("rm in", levels(gg_v$rm_grp))

plot(gg_v, xvar = "lstat", alpha = 0.5) +
  labs(y = st_labs["medv"], x = st_labs["lstat"]) +
  theme(legend.position = "none") +
  scale_color_brewer(palette = "Set3") +
  facet_wrap(~rm_grp)
```

Median values decrease with `lstat` within every `rm` group, but the intercept
shifts upward with more rooms. Smaller homes in low-`lstat` (high-status)
neighborhoods still command high prices.

The complement view --- `medv` vs. `rm`, conditional on `lstat` groups ---
completes the picture:

```{r coplot-lstat, fig.cap="Predicted medv vs. rooms, conditional on lower-status groups."}
lstat_pts <- quantile_pts(rfsrc_Boston$xvar$lstat, groups = 6,
                          intervals = TRUE)
gg_v$lstat_grp <- cut(rfsrc_Boston$xvar$lstat, breaks = lstat_pts)
levels(gg_v$lstat_grp) <- paste("lstat in", levels(gg_v$lstat_grp))

plot(gg_v, xvar = "rm", alpha = 0.5) +
  labs(y = st_labs["medv"], x = st_labs["rm"]) +
  theme(legend.position = "none") +
  scale_color_brewer(palette = "Set3") +
  facet_wrap(~lstat_grp)
```

The `rm` effect is strongest in low-`lstat` tracts (bottom-left panels) and
nearly flat in high-`lstat` tracts, confirming a meaningful interaction.

# Partial Dependence Surface

To visualize the joint partial dependence of `medv` on `lstat` and `rm`, we
compute partial dependence on a grid: 10 values of `rm`, each evaluated at 25
points along `lstat`.

```{r surface-data, cache=TRUE}
rm_grid <- quantile_pts(rfsrc_Boston$xvar$rm, groups = 10)

surface_list <- lapply(rm_grid, function(rm_val) {
  newx <- rfsrc_Boston$xvar
  newx$rm <- rm_val
  pd_rm <- gg_partial_rfsrc(rfsrc_Boston, xvar.names = "lstat", newx = newx)
  df <- pd_rm$continuous
  df$rm <- rm_val
  df
})

surface_df <- bind_rows(surface_list)
```

```{r pd-surface, fig.cap="Partial dependence surface: median home value as a function of lstat and rm. Fill colour is the predicted median value."}
ggplot(surface_df, aes(x = x, y = rm, fill = yhat)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Median Value\n($1000s)") +
  labs(x = "Lower Status (%)", y = "Rooms per Dwelling") +
  theme_bw()
```

The surface confirms the strong interaction: home values are highest when `lstat`
is low and `rm` is high (upper-left corner), dropping steeply along both axes.
The non-planar curvature --- particularly the sharp step near `rm` = 7 ---
demonstrates the kind of complex, non-linear structure that random forests
capture naturally.

# Conclusion

We have walked a full random forest regression analysis with
**randomForestSRC** and **ggRandomForests**, and the pieces line up:

- `gg_error()` showed the OOB error settling well before 500 trees.
- VIMP (`gg_vimp()`) and minimal depth (`max.subtree()`) agreed on the same
  story: `lstat` and `rm` dominate, with a clear gap to everything else.
- `gg_variable()` traced strongly non-linear predictor--response curves, the
  same shapes the raw-data EDA hinted at.
- Partial dependence from `gg_partial_rfsrc()` gave the risk-adjusted version
  of those curves: concave for `lstat`, threshold-like for `rm`.
- Conditioning plots and the partial dependence surface pulled out the `lstat`--`rm`
  interaction, with the room-size effect strongest in high-status tracts.

Notice the pattern in all of this. Each `gg_*()` function returns a tidy
object (often a data frame, sometimes a small list of data frames); the
plotting is a separate step. Use the package's `plot()`
methods when the default figure is what you want, and reach for `ggplot2`
directly when it is not.

# References
