---
title: "Random Hazard Forests with ggRandomForests"
author: "John Ehrlinger"
date: today
format:
  html:
    fig-format: png
    fig-dpi: 96
    toc: true
    toc-depth: 3
    html-math-method: mathjax
    code-fold: false
bibliography: ggRandomForests.bib
vignette: >
  %\VignetteIndexEntry{Random Hazard Forests with ggRandomForests}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# vignette dir under R CMD check, package root under an interactive knit
.fo <- c("_fig_optim.R", file.path("vignettes", "_fig_optim.R"))
if (any(file.exists(.fo))) source(.fo[file.exists(.fo)][1])
knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
options(mc.cores = 1, rf.cores = 1)
```

```{r packages, include=FALSE}
.has_rhf_extractor <- requireNamespace(
  "ggRandomForests",
  quietly = TRUE
) && "gg_rhf" %in% getNamespaceExports("ggRandomForests")
if (.has_rhf_extractor) {
  library(ggRandomForests)
} else if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(
    export_all = FALSE,
    helpers = FALSE,
    attach_testthat = FALSE
  )
} else {
  stop(
    "Install ggRandomForests (or pkgload for dev builds) to render this vignette."
  )
}
```

```{r load-bundle, include=FALSE, cache=FALSE}
.required_bundle_names <- c(
  "data", "fit", "auct_cumulative", "auct_incident", "importance",
  "tune_risk", "tune_iauc", "seed", "settings", "versions"
)
.required_setting_names <- c(
  "formula", "fit", "auct_cumulative", "auct_incident",
  "importance_cache", "importance_time_index", "tune_risk", "tune_iauc"
)
.valid_rhf_bundle <- function(x) {
  if (!is.list(x)) return(FALSE)
  if (!identical(names(x), .required_bundle_names)) return(FALSE)
  checks <- c(
    is.data.frame(x$data),
    inherits(x$fit, "rhf"),
    inherits(x$auct_cumulative, "auct.rhf"),
    inherits(x$auct_incident, "auct.rhf"),
    inherits(x$importance, "importance.rhf"),
    inherits(x$tune_risk, "tune.treesize.rhf"),
    inherits(x$tune_iauc, "tune.treesize.rhf"),
    is.integer(x$seed),
    length(x$seed) == 1L,
    is.list(x$settings),
    identical(names(x$settings), .required_setting_names),
    is.character(x$versions),
    length(x$versions) > 0L
  )
  all(checks)
}

.bundle_paths <- c(
  "rhf_precomputed.rds",
  file.path("vignettes", "rhf_precomputed.rds")
)
bundle <- NULL
for (.bundle_path in .bundle_paths[file.exists(.bundle_paths)]) {
  .candidate_bundle <- tryCatch(
    readRDS(.bundle_path),
    error = function(e) NULL
  )
  if (.valid_rhf_bundle(.candidate_bundle)) {
    bundle <- .candidate_bundle
    break
  }
}
if (is.null(bundle)) {
  stop(
    paste(
      "The RHF vignette needs a valid precomputed bundle.",
      "Run vignettes/precompute_rhf.R from the package root."
    )
  )
}
```

Most survival data begin with one row per subject: a follow-up time, an event
indicator, and covariates measured at baseline. That row cannot show a
covariate changing during follow-up. A Random Hazard Forest (RHF) instead
accepts repeated time intervals for each subject, so a tree can use the
covariate values that belong to each part of the subject's path
[@Ishwaran:RHF:2026; @Ishwaran:RHF:software:2026].

This vignette uses simulated counting-process data to show how
**ggRandomForests** reads an RHF fit. We first make the time-dependent data
structure explicit, then extract case-specific hazard and cumulative hazard
curves. The fitting, performance, importance, and tuning work has already been
run by `vignettes/precompute_rhf.R`. The article evaluates only the inexpensive
**ggRandomForests** extraction and plotting steps.

## What changes with counting-process data?

One subject can contribute several rows. Each row covers an interval
$(\mathit{start}, \mathit{stop}]$, and `event` records whether the event occurs
at that interval's right endpoint. The first simulated subject makes the
layout concrete:

```{r subject-intervals}
rhf_data <- bundle$data
one_subject <- rhf_data[rhf_data$id == rhf_data$id[1L],
                        c("id", "start", "stop", "event", "xtd")]
knitr::kable(one_subject, digits = 3)
```

The intervals do more than divide follow-up into pieces. They define a
covariate path. A time-dependent covariate must be *predictable*: the value
used at time $t$ is known just before the event decision at $t$. In the first
RHF simulation, the actual identity is
`xtd = (x.4 + x.5) * stop`. Because this is a continuous function of time and
the subject's fixed `x.4` and `x.5`, its value at `stop` is also its left-hand
limit there. It is available for the interval ending at that time.

That distinction matters with data collected in practice. A lab value
measured after an interval ends cannot be copied backward into that interval.
At any candidate split time, the subject must be routed using the active
record, not a later measurement. This is the no-lookahead rule: future
covariate values never select an earlier branch. The active-record stitching
and tree routing are behavior of `randomForestRHF::rhf()`, not
**ggRandomForests**. **ggRandomForests** reads the fitted estimates after the
upstream model has applied those rules
[@Ishwaran:RHF:2026; @Ishwaran:RHF:software:2026].

Here is a compact view of the full simulated data set:

```{r data-summary}
records_per_subject <- as.integer(table(rhf_data$id))
events_by_subject <- tapply(rhf_data$event, rhf_data$id, max)
data_summary <- data.frame(
  measure = c(
    "Subjects",
    "Counting-process records",
    "Records per subject",
    "Subjects with an event",
    "xtd range"
  ),
  value = c(
    length(unique(rhf_data$id)),
    nrow(rhf_data),
    sprintf(
      "%d to %d (median %g)",
      min(records_per_subject),
      max(records_per_subject),
      stats::median(records_per_subject)
    ),
    sum(events_by_subject),
    sprintf("%.3f to %.3f", min(rhf_data$xtd), max(rhf_data$xtd))
  )
)
knitr::kable(data_summary, col.names = c("Measure", "Value"))
```

## Where did the fitted objects come from?

The preparation below is shown but not evaluated while the vignette renders.
These are the fixed calls used to create the saved bundle. Keeping the
upstream objects in `rhf_precomputed.rds` makes CRAN builds predictable, while
`vignettes/precompute_rhf.R` lets you reproduce the bundle from the same public
simulation whenever you want to rerun the longer calculations.

```{r preparation, eval=FALSE}
set.seed(20260825L)
sim <- randomForestRHF::hazard.simulation(1)
data <- sim$dta
formula <- stats::as.formula("Surv(id, start, stop, event) ~ .")
with(data, stopifnot(identical(xtd, (x.4 + x.5) * stop)))

fit <- randomForestRHF::rhf(
  formula, data, ntree = 50L, seed = -1L
)
auct_cumulative <- randomForestRHF::auct.rhf(
  fit, marker = "cumhaz", method = "cumulative", verbose = FALSE
)
auct_incident <- randomForestRHF::auct.rhf(
  fit, marker = "hazard", method = "incident",
  riskset = "subject", verbose = FALSE
)
cache <- randomForestRHF::varpro.cache.rhf(
  fit, max.rules.tree = 30L, max.tree = 20L, verbose = FALSE
)
time_index <- unique(as.integer(round(seq.int(
  1L, cache$K, length.out = 5L
))))
importance <- randomForestRHF::importance.rhf(
  fit, cache = cache, time.index = time_index, verbose = FALSE
)
tune_risk <- randomForestRHF::tune.treesize.rhf(
  formula, data, ntree = 20L, perf = "risk", lower = 2L,
  upper = 6L, max.evals = 5L, seed = 20260825L, verbose = FALSE,
  forest = FALSE
)
tune_iauc <- randomForestRHF::tune.iAUC.rhf(
  formula, data, ntree = 20L, lower = 2L, upper = 6L,
  max.evals = 5L, seed = 20260825L, verbose = FALSE,
  forest = FALSE
)
```

The loaded fit can be checked without growing another forest:

```{r fit-summary}
fit_summary <- data.frame(
  measure = c(
    "Object class",
    "Trees",
    "Subjects",
    "Working time points",
    "Saved ensemble source"
  ),
  value = c(
    paste(class(bundle$fit), collapse = ", "),
    bundle$fit$ntree,
    length(bundle$fit$ensemble.id),
    length(bundle$fit$time.interest),
    "out-of-bag hazard and cumulative hazard"
  )
)
knitr::kable(fit_summary, col.names = c("Fit component", "Value"))
```

## How do hazard and cumulative hazard differ?

`gg_rhf()` extracts both estimates into one tidy object, with one row for each
subject and point on the RHF working time grid. This extraction runs live when
the article renders:

```{r extract-rhf}
rhf_curves <- ggRandomForests::gg_rhf(bundle$fit)
head(as.data.frame(rhf_curves))
```

The `source` column is `"oob"` here, so each displayed curve uses the trees
for which that subject was out of bag. We select six subjects only to keep the
lines readable; the extracted object contains all 500.

### Hazard is local

```{r hazard-plot}
case_ids <- c(1, 2, 3, 11, 14, 22)
plot(rhf_curves, idx = case_ids)
```

Hazard is a local event rate. A high point says that events are occurring at a
high instantaneous rate among subjects still at risk around that time. It is
not the probability that a subject has the event at that time, so neither its
height nor its scale should be read as a percentage. RHF estimates this local
quantity directly [@Ishwaran:RHF:2026].

### Why do the hazard curves stop early?

Each line ends where that subject's follow-up ends. This is upstream behavior:
from version 2.0.0, `randomForestRHF::rhf()` returns a pointwise hazard only
where a grid point falls inside one of the subject's supplied `(start, stop]`
records, and `NA` everywhere else, both in the gaps between records and after
the final stop time. Earlier versions carried an estimate across the whole
grid. `gg_rhf()` passes those `NA` values through unchanged, and `plot()` drops
them before drawing, so a curve covers the range the forest was asked about and
no more.

Cumulative hazard is not masked this way. It accumulates the exact overlap
between the working grid and the supplied records, so it stays flat through a
gap rather than going missing, and the panel below draws every grid point. If
you summarize `hazard` yourself, remember to pass `na.rm = TRUE`.

### Cumulative hazard adds the local rates

The same `gg_rhf` object carries `chf`. Set the public `hazard.only` argument
to `FALSE` to draw it:

```{r cumulative-hazard-plot}
plot(rhf_curves, idx = case_ids, hazard.only = FALSE)
```

The cumulative hazard function (CHF) adds hazard over follow-up. It summarizes
accumulated event pressure, but it is unbounded and is not an event
probability. Its scale answers "how much hazard has accumulated?", while the
hazard plot answers "where is the event rate high right now?"

There is one wrinkle in these time-dependent curves. `randomForestRHF` stitches
record-specific ensemble estimates as each subject's active record changes. A
new record's estimated CHF can start below the preceding record's value, so a
stitched case curve may contain a downward step. That step does not mean a
negative hazard or that past event pressure disappeared. It marks the upstream
model's transition to a different active-record estimate
[@Ishwaran:RHF:software:2026].

## How does discrimination change over time?

Survival AUC needs a definition of a case and a control at each time.
Different definitions answer different questions, and each uses the marker
that matches its question. This section covers the incident/dynamic
definition. The saved AUC calculation is supplied to `gg_auct()`; only the
extraction and plotting run during this render.

### Incident/dynamic AUC uses hazard

The question here is whether a subject who experiences the event near time $t$
ranks above a subject in the relevant risk set at $t$. This is the
incident/dynamic target. It uses the local hazard marker, named `"haz"` in
`gg_auct()` and `"hazard"` upstream.

```{r auct-incident}
auct_incident <- gg_auct(
  bundle$fit,
  marker = "haz",
  auct_fit = bundle$auct_incident
)
plot(auct_incident)
```

The saved curve has a Uno iAUC of
`r sprintf("%.3f", attr(auct_incident, "iauc")$uno)`. Its finite AUC values
range from `r sprintf("%.3f", min(auct_incident$auc, na.rm = TRUE))` to
`r sprintf("%.3f", max(auct_incident$auc, na.rm = TRUE))` over this fitted
time grid. AUC is a ranking probability on a 0 to 1 scale, with 0.5 shown as
the chance reference. The retained calculation did not use a bootstrap, so the
plot has no confidence ribbon.

### Why cumulative/dynamic AUC is not shown here

The companion definition is cumulative/dynamic AUC, which asks whether a
subject who has experienced the event by a horizon ranks above a subject who
remains event-free at that horizon. It ranks accumulated risk through a
horizon rather than local failures within a risk set, so it estimates a
different target, not a better or worse version of the same one. Its matching
marker is cumulative hazard.

We have left it out for now. From `randomForestRHF` 2.0.0, the in-sample
cumulative hazard is held flat once a subject's supplied records end, which
`?rhf` documents. That is reasonable for a fitted curve, but it means the
cumulative hazard at a fixed grid time reflects how long each subject was
observed as well as how much risk they carried, and the cumulative/dynamic
comparison puts subjects who have already failed against subjects still under
follow-up. On this simulated data that is enough to push the curve below the
chance line. The behavior is upstream rather than in `gg_auct()`, we have
reported it, and the section will come back once it is settled.

## Which variables matter, and when?

The retained `importance.rhf` object is the starting point here. Its expensive
upstream work was performed once: `varpro.cache.rhf()` collected the rule and
near-miss information, then `importance.rhf()` evaluated five selected time
windows. Supplying that result avoids repeating either calculation.

```{r rhf-importance, fig.width=8, fig.height=5.25}
rhf_priority <- gg_rhf_importance(
  bundle$fit,
  importance_fit = bundle$importance
)
plot(rhf_priority) +
  ggplot2::theme(plot.margin = ggplot2::margin(l = 80))
```

RHF priority is a time-local rule-release contrast. Within a window, it
compares the log integrated-hazard working response for a rule with its
near-miss set and asks how much that fitted response changes when rules
involving a variable are released. Larger nonnegative values in this supplied
object mean a larger contrast. They do not say that a covariate raises or
lowers hazard, and they are not z-scores, p-values, or automatic variable
selection thresholds.

The dot size and color show the same priority value, with a 99th-percentile
cap applied to the display only. The extracted values do not change. In this
simulation, `xtd` has the largest early-window value, about
`r sprintf("%.2f", max(rhf_priority$priority, na.rm = TRUE))`, while `x.1`
has the largest value in several later windows. The number at risk falls from
`r max(rhf_priority$n_risk)` to `r min(rhf_priority$n_risk)` across the five
displayed windows, so the late-window comparisons use less information. One
late `x.4` result is unavailable in the upstream object and appears without a
point. For work you plan to revisit, retain both the upstream cache and the
`importance.rhf` result; supply the latter to `gg_rhf_importance()` for each
display.

## Which tree size did the upstream searches select?

The two retained `tune.treesize.rhf` objects contain the evaluated paths and
their selected sizes. `gg_tune_rhf()` preserves the upstream evaluation order,
so the connecting line follows the order in which candidates were evaluated,
not a newly sorted sequence. The emphasized point marks the size selected by
the upstream search.

### OOB risk tuning

```{r tune-risk}
risk_tuning <- gg_tune_rhf(bundle$tune_risk)
risk_tuning
plot(risk_tuning)
```

The risk search evaluated tree sizes
`r paste(risk_tuning$treesize, collapse = ", ")` in that order and selected
size `r risk_tuning$treesize[risk_tuning$selected]`, where the saved mean OOB
risk was smallest. The retained risk path has no standard-error field.
`gg_tune_rhf()` sets its `se` values to missing and does not fabricate
uncertainty, so a risk plot has no ribbon.

### OOB iAUC tuning

```{r tune-iauc}
iauc_tuning <- gg_tune_rhf(bundle$tune_iauc)
iauc_tuning
plot(iauc_tuning)
```

The iAUC search evaluated tree sizes
`r paste(iauc_tuning$treesize, collapse = ", ")` and selected size
`r iauc_tuning$treesize[iauc_tuning$selected]`, where OOB iAUC was largest.
An iAUC ribbon is drawn only when finite upstream standard errors are present.
This saved path has `r sum(is.finite(iauc_tuning$se))` finite standard errors,
so its plot has no ribbon. Retaining each tuning result preserves the actual
candidate path, search order, and selected point without rerunning the fits.

## What does each extractor return?

All four families keep extraction separate from plotting. The intermediate
object is a data frame with an added class, so you can inspect it before
calling `plot()`.

```{r extractor-support}
support <- data.frame(
  extractor = c(
    "gg_rhf()",
    "gg_auct()",
    "gg_rhf_importance()",
    "gg_tune_rhf()"
  ),
  upstream_input = c(
    "rhf fit",
    "rhf fit plus an optional auct.rhf object",
    "rhf fit plus an optional importance.rhf object",
    "tune.treesize.rhf object"
  ),
  returned_class = c(
    "gg_rhf, data.frame",
    "gg_auct, data.frame",
    "gg_rhf_importance, data.frame",
    "gg_tune_rhf, data.frame"
  ),
  scale = c(
    paste(
      "Hazard is a local event rate; cumulative hazard is accumulated",
      "event pressure. Neither is a probability."
    ),
    paste(
      "AUC(t) and iAUC are ranking measures from 0 to 1; interpretation",
      "depends on the case definition and marker."
    ),
    paste(
      "Time-local rule-release priority; larger values mean a larger",
      "working-response contrast, not statistical significance."
    ),
    paste(
      "OOB risk is minimized or OOB iAUC is maximized; the selected",
      "tree size is marked."
    )
  ),
  check.names = FALSE
)
knitr::kable(
  support,
  col.names = c(
    "Extractor", "Upstream input class or object",
    "Returned class", "Output scale and interpretation"
  )
)
```

## How can this analysis be reproduced?

The bundle records the public simulation, seed, calculation settings, and
package versions used for this render. These are the saved versions:

```{r bundle-versions}
version_table <- data.frame(
  package = names(bundle$versions),
  version = unname(bundle$versions),
  row.names = NULL
)
knitr::kable(
  version_table,
  col.names = c("Software", "Saved version")
)
```

Run `Rscript vignettes/precompute_rhf.R` from the package root to regenerate
all ten bundle components. That script repeats the sole longitudinal
simulation and uses `bundle$seed` and the settings shown earlier. The exact
numerical result can still change when the recorded package versions change.

For routine use, retain the fitted forest and each expensive upstream result:
the two `auct.rhf` calculations, the importance cache and result, and both
tuning objects. Supplying those objects is the default workflow for a report
you expect to render more than once. It separates a long model calculation
from a short, repeatable extraction and plot.

## Further reading

The RHF method, including its counting-process construction and hazard-based
forest, is described by Ishwaran, Hsich, Kogalur, and Lee
[@Ishwaran:RHF:2026]. The `randomForestRHF` software reference documents the
R implementation and the upstream objects used throughout this article
[@Ishwaran:RHF:software:2026].
