---
title: "Random Hazard Forests with ggRandomForests"
author: "John Ehrlinger"
date: today
format:
  html:
    fig-format: png
    fig-dpi: 96
    toc: true
    toc-depth: 3
    html-math-method: mathjax
    code-fold: false
bibliography: ggRandomForests.bib
vignette: >
  %\VignetteIndexEntry{Random Hazard Forests with ggRandomForests}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# vignette dir under R CMD check, package root under an interactive knit
.fo <- c("_fig_optim.R", file.path("vignettes", "_fig_optim.R"))
if (any(file.exists(.fo))) source(.fo[file.exists(.fo)][1])
knitr::opts_chunk$set(
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
options(mc.cores = 1, rf.cores = 1)
```

```{r packages, include=FALSE}
.has_rhf_extractor <- requireNamespace(
  "ggRandomForests",
  quietly = TRUE
) && "gg_rhf" %in% getNamespaceExports("ggRandomForests")
if (.has_rhf_extractor) {
  library(ggRandomForests)
} else if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(
    export_all = FALSE,
    helpers = FALSE,
    attach_testthat = FALSE
  )
} else {
  stop(
    "Install ggRandomForests (or pkgload for dev builds) to render this vignette."
  )
}
```

```{r load-bundle, include=FALSE, cache=FALSE}
.required_bundle_names <- c(
  "data", "fit", "auct_cumulative", "auct_incident", "importance",
  "tune_risk", "tune_iauc", "seed", "settings", "versions"
)
.required_setting_names <- c(
  "formula", "fit", "auct_cumulative", "auct_incident",
  "importance_cache", "importance_time_index", "tune_risk", "tune_iauc"
)
.valid_rhf_bundle <- function(x) {
  if (!is.list(x)) return(FALSE)
  if (!identical(names(x), .required_bundle_names)) return(FALSE)
  checks <- c(
    is.data.frame(x$data),
    inherits(x$fit, "rhf"),
    inherits(x$auct_cumulative, "auct.rhf"),
    inherits(x$auct_incident, "auct.rhf"),
    inherits(x$importance, "importance.rhf"),
    inherits(x$tune_risk, "tune.treesize.rhf"),
    inherits(x$tune_iauc, "tune.treesize.rhf"),
    is.integer(x$seed),
    length(x$seed) == 1L,
    is.list(x$settings),
    identical(names(x$settings), .required_setting_names),
    is.character(x$versions),
    length(x$versions) > 0L
  )
  all(checks)
}

.bundle_paths <- c(
  "rhf_precomputed.rds",
  file.path("vignettes", "rhf_precomputed.rds")
)
bundle <- NULL
for (.bundle_path in .bundle_paths[file.exists(.bundle_paths)]) {
  .candidate_bundle <- tryCatch(
    readRDS(.bundle_path),
    error = function(e) NULL
  )
  if (.valid_rhf_bundle(.candidate_bundle)) {
    bundle <- .candidate_bundle
    break
  }
}
if (is.null(bundle)) {
  stop(
    paste(
      "The RHF vignette needs a valid precomputed bundle.",
      "Run vignettes/precompute_rhf.R from the package root."
    )
  )
}
```

Most survival data begin with one row per subject: a follow-up time, an event
indicator, and covariates measured at baseline. That row cannot show a
covariate changing during follow-up. A Random Hazard Forest (RHF) instead
accepts repeated time intervals for each subject, so a tree can use the
covariate values that belong to each part of the subject's path
[@Ishwaran:RHF:2026; @Ishwaran:RHF:software:2026].

This vignette uses simulated counting-process data to show how
**ggRandomForests** reads an RHF fit. We first make the time-dependent data
structure explicit, then extract case-specific hazard and cumulative hazard
curves. The fitting, performance, importance, and tuning work has already been
run by `vignettes/precompute_rhf.R`. The article evaluates only the inexpensive
**ggRandomForests** extraction and plotting steps.

## What changes with counting-process data?

One subject can contribute several rows. Each row covers an interval
$(\mathit{start}, \mathit{stop}]$, and `event` records whether the event occurs
at that interval's right endpoint. The first simulated subject makes the
layout concrete:

```{r subject-intervals}
rhf_data <- bundle$data
one_subject <- rhf_data[rhf_data$id == rhf_data$id[1L],
                        c("id", "start", "stop", "event", "xtd")]
knitr::kable(one_subject, digits = 3)
```

The intervals do more than divide follow-up into pieces. They define a
covariate path. A time-dependent covariate must be *predictable*: the value
used at time $t$ is known just before the event decision at $t$. In the first
RHF simulation, the actual identity is
`xtd = (x.4 + x.5) * stop`. Because this is a continuous function of time and
the subject's fixed `x.4` and `x.5`, its value at `stop` is also its left-hand
limit there. It is available for the interval ending at that time.

That distinction matters with data collected in practice. A lab value
measured after an interval ends cannot be copied backward into that interval.
At any candidate split time, the subject must be routed using the active
record, not a later measurement. This is the no-lookahead rule: future
covariate values never select an earlier branch. The active-record stitching
and tree routing are behavior of `randomForestRHF::rhf()`, not
**ggRandomForests**. **ggRandomForests** reads the fitted estimates after the
upstream model has applied those rules
[@Ishwaran:RHF:2026; @Ishwaran:RHF:software:2026].

Here is a compact view of the full simulated data set:

```{r data-summary}
records_per_subject <- as.integer(table(rhf_data$id))
events_by_subject <- tapply(rhf_data$event, rhf_data$id, max)
data_summary <- data.frame(
  measure = c(
    "Subjects",
    "Counting-process records",
    "Records per subject",
    "Subjects with an event",
    "xtd range"
  ),
  value = c(
    length(unique(rhf_data$id)),
    nrow(rhf_data),
    sprintf(
      "%d to %d (median %g)",
      min(records_per_subject),
      max(records_per_subject),
      stats::median(records_per_subject)
    ),
    sum(events_by_subject),
    sprintf("%.3f to %.3f", min(rhf_data$xtd), max(rhf_data$xtd))
  )
)
knitr::kable(data_summary, col.names = c("Measure", "Value"))
```

## Where did the fitted objects come from?

The preparation below is shown but not evaluated while the vignette renders.
These are the fixed calls used to create the saved bundle. Keeping the
upstream objects in `rhf_precomputed.rds` makes CRAN builds predictable, while
`vignettes/precompute_rhf.R` lets you reproduce the bundle from the same public
simulation whenever you want to rerun the longer calculations.

```{r preparation, eval=FALSE}
set.seed(20260825L)
sim <- randomForestRHF::hazard.simulation(1)
data <- sim$dta
formula <- stats::as.formula("Surv(id, start, stop, event) ~ .")
with(data, stopifnot(identical(xtd, (x.4 + x.5) * stop)))

fit <- randomForestRHF::rhf(
  formula, data, ntree = 50L, seed = -1L
)
auct_cumulative <- randomForestRHF::auct.rhf(
  fit, marker = "cumhaz", method = "cumulative", verbose = FALSE
)
auct_incident <- randomForestRHF::auct.rhf(
  fit, marker = "hazard", method = "incident",
  riskset = "subject", verbose = FALSE
)
cache <- randomForestRHF::varpro.cache.rhf(
  fit, max.rules.tree = 30L, max.tree = 20L, verbose = FALSE
)
time_index <- unique(as.integer(round(seq.int(
  1L, cache$K, length.out = 5L
))))
importance <- randomForestRHF::importance.rhf(
  fit, cache = cache, time.index = time_index, verbose = FALSE
)
tune_risk <- randomForestRHF::tune.treesize.rhf(
  formula, data, ntree = 20L, perf = "risk", lower = 2L,
  upper = 6L, max.evals = 5L, seed = 20260825L, verbose = FALSE,
  forest = FALSE
)
tune_iauc <- randomForestRHF::tune.iAUC.rhf(
  formula, data, ntree = 20L, lower = 2L, upper = 6L,
  max.evals = 5L, seed = 20260825L, verbose = FALSE,
  forest = FALSE
)
```

The loaded fit can be checked without growing another forest:

```{r fit-summary}
fit_summary <- data.frame(
  measure = c(
    "Object class",
    "Trees",
    "Subjects",
    "Working time points",
    "Saved ensemble source"
  ),
  value = c(
    paste(class(bundle$fit), collapse = ", "),
    bundle$fit$ntree,
    length(bundle$fit$ensemble.id),
    length(bundle$fit$time.interest),
    "out-of-bag hazard and cumulative hazard"
  )
)
knitr::kable(fit_summary, col.names = c("Fit component", "Value"))
```

## How do hazard and cumulative hazard differ?

`gg_rhf()` extracts both estimates into one tidy object, with one row for each
subject and point on the RHF working time grid. This extraction runs live when
the article renders:

```{r extract-rhf}
rhf_curves <- ggRandomForests::gg_rhf(bundle$fit)
head(as.data.frame(rhf_curves))
```

The `source` column is `"oob"` here, so each displayed curve uses the trees
for which that subject was out of bag. We select six subjects only to keep the
lines readable; the extracted object contains all 500.

### Hazard is local

```{r hazard-plot}
case_ids <- c(1, 2, 3, 11, 14, 22)
plot(rhf_curves, idx = case_ids)
```

Hazard is a local event rate. A high point says that events are occurring at a
high instantaneous rate among subjects still at risk around that time. It is
not the probability that a subject has the event at that time, so neither its
height nor its scale should be read as a percentage. RHF estimates this local
quantity directly [@Ishwaran:RHF:2026].

### Cumulative hazard adds the local rates

The same `gg_rhf` object carries `chf`. Set the public `hazard.only` argument
to `FALSE` to draw it:

```{r cumulative-hazard-plot}
plot(rhf_curves, idx = case_ids, hazard.only = FALSE)
```

The cumulative hazard function (CHF) adds hazard over follow-up. It summarizes
accumulated event pressure, but it is unbounded and is not an event
probability. Its scale answers "how much hazard has accumulated?", while the
hazard plot answers "where is the event rate high right now?"

There is one wrinkle in these time-dependent curves. `randomForestRHF` stitches
record-specific ensemble estimates as each subject's active record changes. A
new record's estimated CHF can start below the preceding record's value, so a
stitched case curve may contain a downward step. That step does not mean a
negative hazard or that past event pressure disappeared. It marks the upstream
model's transition to a different active-record estimate
[@Ishwaran:RHF:software:2026].
