---
title: "Exploring variable importance with varPro"
author: "John Ehrlinger"
date: today
format:
  html:
    fig-format: png
    fig-dpi: 96
    toc: true
    toc-depth: 3
    html-math-method: mathjax
    code-fold: false
bibliography: ggRandomForests.bib
vignette: >
  %\VignetteIndexEntry{Exploring variable importance with varPro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  cache      = TRUE,
  cache.path = "varpro_cache/"
)
options(mc.cores = 1, rf.cores = 1)
```

```{r libs}
library(ggplot2)

# Match the pattern used by the regression/survival vignettes: try the
# installed package first, fall back to pkgload::load_all() for the
# R CMD check vignette rebuild where the package isn't yet on
# .libPaths(). All varPro calls below are ::-qualified, so no
# library(varPro) is needed.
if (requireNamespace("ggRandomForests", quietly = TRUE)) {
  library(ggRandomForests)
} else if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(export_all = FALSE, helpers = FALSE,
                    attach_testthat = FALSE)
} else {
  stop("Install ggRandomForests (or pkgload for dev builds) to render this vignette.")
}
```

```{r precomputed, include=FALSE, cache=FALSE}
# The three gg_partial_varpro() calls and the Boston beta.varpro() fit are
# the only expensive parts of this vignette (~34s combined). They are
# precomputed by precompute_varpro.R and loaded here so the R CMD check
# vignette rebuild stays fast. If the file is absent (e.g. a fresh clone
# that hasn't run the script), each chunk falls back to live computation,
# so the vignette is reproducible either way.
.vp <- tryCatch({
  # R CMD check rebuilds the vignette in a copy whose working directory is the
  # vignette dir; an interactive knit from the package root sees vignettes/
  # instead. Try both locations, and fall back to live computation below if
  # neither is present or readRDS() fails.
  .cand <- c("varpro_precomputed.rds",
             file.path("vignettes", "varpro_precomputed.rds"))
  .hit <- .cand[file.exists(.cand)]
  if (length(.hit)) readRDS(.hit[[1]]) else list()
}, error = function(e) list())   # missing or unreadable -> live fallback below
```

# What varPro is

Random-forest variable importance has, for two decades, mostly meant
*permutation importance*: shuffle a column, measure the drop in OOB
accuracy, repeat. It works, but the score is computed against
artificially modified data that never appeared in the training set, and
when predictors are correlated the shuffled column is an implausible
counterfactual, not a neutral baseline. Knockoff-style methods clean up
some of that, but they introduce their own synthetic features and rely on
explicit distributional assumptions about the predictors.

varPro [@Lu2024varpro] takes a different path: one grounded entirely in
*observed* data. Think of each decision tree as a long chain of "if/then"
clauses. varPro harvests those clauses as *rules*, and for each rule it
identifies a specific region of the predictor space where a handful of
variables jointly constrain the response. To measure one variable's
contribution, it compares a *local estimator* (the response summary
restricted to that rule's region) against a "released" estimator where
the constraint on the tested variable has been lifted. That comparison,
averaged over rules and trees, is the variable's importance z-score. Every
observation used in the contrast is real training data; nothing is
permuted, nothing is synthesised.

The "release" metaphor is the right one. Imagine holding a rope at several
points along its length (the rule constraints). Releasing your grip at one
point (while keeping the others in place) tells you how taut that
section was. Variables that go slack when released were carrying load;
variables that stay taut were redundant given the rest.

Operationally, varPro's *guided splitting* steers tree growth so that
rules concentrate in informative regions, and the subsequent *rule
harvesting* step selects the rules that survived the importance
pre-filter. Variables below a noise threshold are dropped before
reporting (the importance z is pre-filtered), so the final ranking
contains only variables that earned a seat at the table
(see the [varPro tools reference](https://www.varprotools.org/articles/getstarted.html)
for worked implementation details). Split-weights encode each variable's
relative importance at each tree node, propagating that signal through the
release-rule contrasts and into the partial-dependence curves.

Cross-validation via `cv.varpro()` returns three views of the
importance ranking: `$imp` (the default, point-estimate ranking),
`$imp.conserve` (one-SD rule: only variables whose lower confidence
bound is positive), and `$imp.liberal` (any variable with a positive
point estimate). The conservative cut is the right default when you want
to hand a short list to a collaborator; the liberal cut is better for
exploratory screening when you'd rather miss nothing.

For classification, the importance decomposes into an `$unconditional`
score (averaged over all classes) and per-class `$conditional.z` scores.
When a variable separates *one* class from the others but is uninformative
overall, the unconditional z can be near zero while a single conditional z
is large; the conditional view catches that. One-hot consolidation via
`get.orgvimp()` and `get.topvars()` maps the per-level scores back to the
original factor, so downstream reporting stays on familiar ground.

That core machinery feeds six ggRandomForests wrappers. **`gg_varpro()`**
summarises the per-tree importance distribution. **`gg_beta_varpro()`**
refines those release-rule contrasts with a per-rule lasso. **`gg_partial_varpro()`**
turns the release machinery into partial-dependence curves.
**`gg_udependent()`** reads cross-variable dependency off a `uvarpro()`
fit. **`gg_isopro()`** scores observations for anomaly using an
isolation-forest variant. **`gg_ivarpro()`** computes per-observation
local importance.

This vignette walks all six wrappers on three worked examples: a
regression problem (Boston housing), a classification problem (iris,
binary and multi-class), and a survival problem (PBC). The closing
section is a one-page reference matrix mapping each wrapper to the
forest families it supports.

# Regression: Boston housing

We start with the Boston housing data: 506 census tracts, 13 numeric
predictors, median home value as the response.

```{r boston-fit}
data("Boston", package = "MASS")
set.seed(20260527L)
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
v_boston <- if (is.null(.vp$v_boston)) {
  varPro::varpro(medv ~ ., data = Boston, ntree = 50)
} else {
  .vp$v_boston
}
v_boston
```

`varpro()` builds the importance machinery (release rules, per-tree
scores, the splitweight vector) once. Every subsequent figure in this
section reads off that single fit. The formula interface is intentional:
varPro is model-independent in the sense that it doesn't assume a linear
or parametric relationship between predictors and response, but it does
require you to name a response so the guided splitting has something to
optimise toward.

## Per-tree importance with `gg_varpro()`

The first question varPro answers is the same one permutation importance
answers: which variables matter, ranked. `gg_varpro()` extracts the
per-tree importance scores and draws their distribution as a horizontal
boxplot, one row per variable, sorted by median z-score. Variables above
a configurable cutoff (default 0.79) are highlighted.

```{r boston-gg-varpro}
plot(gg_varpro(v_boston))
```

The narrow boxes near the top are the variables varPro is confident
about: every tree agrees they matter. Wide boxes that straddle the
cutoff line are the ones to look at twice; the forest disagrees with
itself. That disagreement isn't noise to suppress: it can mean the
variable matters in some rule regions but not others, which is exactly
the kind of structured heterogeneity that model-independent methods are
built to detect. Variables that fall entirely below the cutoff were
pre-filtered by the importance z threshold; they're gone before the plot
is drawn.

## Partial dependence with `gg_partial_varpro()`

The ranking view tells you *which* variables matter. Partial dependence
asks the next question: *how* does the response change with a variable,
holding the others fixed? `gg_partial_varpro()` wraps
`varPro::partialpro()` and returns a tidy frame of parametric,
non-parametric, and causal partial-dependence curves.

```{r boston-gg-partial-varpro}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
gg_pd <- if (is.null(.vp$pd_boston)) {
  gg_partial_varpro(object = v_boston)
} else {
  .vp$pd_boston
}
plot(gg_pd)
```

Each panel is a single predictor. The three curves correspond to the
three estimators varPro carries (parametric, non-parametric, and causal);
read them as a sensitivity analysis. When all three agree, you have a
stable signal; when the causal curve diverges from the others, that's a
hint that the variable's observed relationship with the response may be
partly driven by its correlation with other predictors in the rule
regions, not by a direct effect.

Because `partialpro()` operates within the *local* neighborhoods defined
by the release rules, these curves reflect what happens in the part of
the predictor space the forest actually visited, not a global
extrapolation from a fitted model. That is both a strength (no
out-of-distribution extrapolation) and a limit (sparse regions of the
predictor space will have wider effective intervals, even if the plot
doesn't always show them explicitly).

## Per-rule lasso refinement with `gg_beta_varpro()`

`gg_varpro()`'s ranking is built from the release-rule contrast.
`gg_beta_varpro()` re-asks the question variable by variable inside each
rule: it fits a one-predictor lasso of the response on the released
variable, restricted to the OOB observations in that rule's region. The
fitted coefficient (call it a "local β") captures the variable's
linear effect *within that rule's neighborhood*, not globally. Aggregating
`mean(|β|)` across rules gives one number per variable: a
regression-coefficient-flavoured importance, not a VIMP score, and not a
global slope.

That distinction matters in practice. A variable with a strong nonlinear
global relationship may have locally small β values inside any single
rule (the local-standardisation step within each rule normalises the
scale), but many rules will fire on it, so the aggregated mean is still
large. Conversely, a variable with a nearly linear global effect will
concentrate most of its weight in a handful of rules, and the
between-rule variability in β will be low.

Because `beta.varpro()` is expensive (a `glmnet` per rule), the wrapper
accepts a pre-computed `beta_fit` so you can iterate on selection,
cutoff, or class choice without re-fitting.

```{r boston-beta-fit, cache=TRUE}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
b_boston <- if (is.null(.vp$b_boston)) {
  varPro::beta.varpro(v_boston)
} else {
  .vp$b_boston
}
```

```{r boston-gg-beta-varpro}
plot(gg_beta_varpro(v_boston, beta_fit = b_boston))
```

Compare against `gg_varpro()` above. Disagreement is the diagnostic
signal: a variable that ranks high here but low there is one whose
local linear effect inside many rules is real even when the release
contrast is modest.

## Cross-variable dependency with `gg_udependent()`

The three views so far take one variable at a time. `gg_udependent()`
reads cross-variable structure off a `uvarpro()` fit and draws the
result as a network: nodes are variables, edges are dependencies above a
configurable threshold. The visual is built with `ggraph`, which is in
`Suggests` rather than `Imports`; install it if you want this view.

```{r boston-uvarpro, cache=TRUE}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
u_boston <- if (is.null(.vp$u_boston)) {
  varPro::uvarpro(Boston[, setdiff(names(Boston), "medv")], ntree = 50)
} else {
  .vp$u_boston
}
```

```{r boston-gg-udependent, eval = requireNamespace("ggraph", quietly = TRUE)}
plot(gg_udependent(u_boston))
```

Clusters of mutually-connected variables are worth checking for
redundancy: they may be several views of the same underlying quantity.
`uvarpro()` operates on the predictor matrix alone; there is no response
in the fit, which is what makes the network genuinely unsupervised.
Variables that cluster here are correlated in feature space regardless of
whether any of them matters for prediction. That information is
complementary to the ranking from `gg_varpro()`: a variable can be
important for prediction and still sit in a tight cluster with a
near-duplicate; in that situation, model parsimony may favour dropping
one of the cluster members without losing much predictive accuracy.

## Anomaly scoring with `gg_isopro()`

Variable importance is one axis; *observation* outlierness is another.
`gg_isopro()` wraps `varPro::isopro()` (an isolation-forest variant
that scores how anomalous each training row looks) and renders the
result as a ranked elbow plus a density of the scores. The score is
on `[0, 1]`; the wrapper's convention is "higher = more anomalous"
(opposite of varPro's native polarity; the wrapper flips it for
consistency).

```{r boston-isopro, cache=TRUE}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
iso_boston <- if (is.null(.vp$iso_boston)) {
  varPro::isopro(data = Boston[, setdiff(names(Boston), "medv")],
                 method = "rnd", sampsize = 256, ntree = 50)
} else {
  .vp$iso_boston
}
```

```{r boston-gg-isopro}
plot(gg_isopro(iso_boston))
```

The elbow flags rows that diverge from the bulk. Pair with the
domain: anomalous in feature space is not the same as wrong, but it's
often where the most interesting cases live. In a regression context,
a high-scoring row is one that the isolation forest isolated quickly
because its feature combination is rare; it may or may not be an outlier
in the response. Anomaly scoring and residual analysis answer different
questions, and it's worth doing both.

Note that `isopro()` scores are not calibrated to a universal scale: a
score of 0.7 in one dataset is not comparable to 0.7 in another. What
matters is the relative ordering within a dataset and the shape of the
elbow: a sharp kink at a small number of observations is a cleaner
signal than a gradual slope that never levels off.

## Local importance with `gg_ivarpro()`

The wrappers so far aggregate across observations. `gg_ivarpro()` does
the opposite: it returns one value per (observation, variable) pair,
capturing how much variable *v* contributed to predicting observation
*i*. The aggregate view is a distribution of those local importances
per variable; the per-observation view is a horizontal bar of one row's
local importances across variables.

`ivarpro()` is the most expensive call in varPro, so the wrapper
accepts a pre-computed `ivarpro_fit` for reuse across views.

```{r boston-ivarpro-fit, cache=TRUE}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
iv_boston <- if (is.null(.vp$iv_boston)) {
  varPro::ivarpro(v_boston)
} else {
  .vp$iv_boston
}
```

```{r boston-gg-ivarpro-distribution}
plot(gg_ivarpro(v_boston, ivarpro_fit = iv_boston))
```

```{r boston-gg-ivarpro-which-obs}
plot(gg_ivarpro(v_boston, ivarpro_fit = iv_boston, which_obs = 1L))
```

The distribution view tells you which variables drive predictions
*across* observations. The per-observation view answers the same
question for a specific case, useful for explaining one prediction
back to whoever asked.

The per-observation importance is computed by replaying the release-rule
contrasts restricted to the rules that fire for that observation. It
is not a Shapley value, but it shares the local-attribution spirit: each
bar shows how much that variable's release shifted the local estimator
for this specific row, averaged over the rules that covered it. Cases
that are well-represented in the training data (dense rule coverage) will
have sharper attribution than cases in sparse regions of the predictor
space.

# Classification: iris

Iris is a small data set (150 rows, four predictors, three response
classes), and that's a feature here, not a flaw: every figure renders in
under a second, and the structure is well-understood enough that any
strange behaviour stands out. It is also a good stress-test for the
conditional importance path: petal length and petal width separate
*setosa* from everything else very cleanly, but the *versicolor*/*virginica*
boundary is much softer. A method that only reports unconditional
importance would lump both cases together; the conditional decomposition
should show the asymmetry.

Two fits: a binary problem (drop *setosa*, positive class = *virginica*)
and the full three-class problem.

```{r iris-fit-binary}
iris_binary <- iris[iris$Species != "setosa", ]
iris_binary$Species <- droplevels(iris_binary$Species)
set.seed(20260527L)
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
v_iris_binary <- if (is.null(.vp$v_iris_binary)) {
  varPro::varpro(Species ~ ., data = iris_binary, ntree = 50)
} else {
  .vp$v_iris_binary
}
```

```{r iris-fit-multi}
set.seed(20260527L)
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
v_iris_multi <- if (is.null(.vp$v_iris_multi)) {
  varPro::varpro(Species ~ ., data = iris, ntree = 50)
} else {
  .vp$v_iris_multi
}
```

## Class-conditional importance with `gg_varpro(conditional = TRUE)`

For a classification forest, `gg_varpro()` can split the importance
view into one facet per class. Variables keep the unconditional sort
order so rows line up across facets; read along a row to see which
class a variable is informative for.

The conditional importance uses `$conditional.z` from the varPro fit:
per-class release-rule contrasts computed within the same rule regions
as the unconditional score, but with the local estimator replaced by a
per-class probability. A variable that separates one class from the rest
will have a high `$conditional.z` for that class and a near-zero or
even negative value for the others. The unconditional score in
`$unconditional` is the average, and can mask class-specific effects,
which is why the `conditional = TRUE` faceted view exists.

```{r iris-multi-gg-varpro-conditional}
plot(gg_varpro(v_iris_multi, conditional = TRUE))
```

## Partial dependence: `gg_partial_varpro()` on classification

On a classification fit `gg_partial_varpro()` defaults to the
**probability scale** (`scale = "auto"` resolves to `"prob"`): each
predictor's curve is the predicted probability of the *target class* —
by default the last factor level (here *virginica*), selectable with
`target =`. `varPro::partialpro()` works internally on the log-odds of
that class; the wrapper back-transforms each observation to a
probability *before* averaging, so the curve is the mean predicted
probability (not the probability of the mean log-odds). On this bounded
$[0, 1]$ scale only the parametric and non-parametric curves are shown —
the `causal` contrast is a log odds-ratio, not a level, so it cannot
share the probability axis (use `scale = "logodds"` to see it).
`scale = "odds"` and `"logodds"` give the same relationship on the odds
and log-odds scales.

```{r iris-multi-gg-partial-varpro}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
gg_pd_iris <- if (is.null(.vp$pd_iris_multi)) {
  gg_partial_varpro(object = v_iris_multi)
} else {
  .vp$pd_iris_multi
}
plot(gg_pd_iris)
```

Read each panel as: "as predictor X changes from low to high, how does
the probability of the target class shift?" Patterns that match the
underlying biology (e.g. petal length separating *virginica* from the
others) act as a sanity check on the forest.

## Per-class lasso refinement with `gg_beta_varpro()`

On a classification fit, `gg_beta_varpro()` returns one row per
(variable, class) pair. For a binary fit, `which_class = NULL` defaults
to the last factor level (the positive class), so the headline view
is a single panel of that class.[^mortality]

[^mortality]: The same code pattern applies to clinical binary outcomes
such as 30-day mortality: drop the negative class, set the event class
as the last factor level, and read the figure for the event panel.

```{r iris-binary-beta-fit, cache=TRUE}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
b_iris_binary <- if (is.null(.vp$b_iris_binary)) {
  varPro::beta.varpro(v_iris_binary)
} else {
  .vp$b_iris_binary
}
```

```{r iris-binary-gg-beta-varpro}
plot(gg_beta_varpro(v_iris_binary, beta_fit = b_iris_binary))
```

For a multi-class fit, the default view is faceted by class with each
class sharing the row order set by the unified ranking, same trick as
`gg_varpro(conditional = TRUE)`.

```{r iris-multi-beta-fit, cache=TRUE}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
b_iris_multi <- if (is.null(.vp$b_iris_multi)) {
  varPro::beta.varpro(v_iris_multi)
} else {
  .vp$b_iris_multi
}
```

```{r iris-multi-gg-beta-varpro}
plot(gg_beta_varpro(v_iris_multi, beta_fit = b_iris_multi))
```

`which_class = "<level>"` collapses the faceted view to a single class
when you want it.

# Survival: PBC

Survival is the family where the varPro toolchain shows its limits,
and being explicit about those limits is more useful than pretending they
don't exist. The forest-fitting and the family-agnostic wrappers all
work; the lasso-refined and individual-importance wrappers don't, because
the underlying `varPro::beta.varpro()` and `varPro::ivarpro()` calls
don't yet extend to right-censored outcomes. A per-rule local lasso for a
censored response requires a local partial-likelihood or Nelson-Aalen
estimator in place of the regression/classification local estimator;
the design work for that is tracked but not yet landed.

What does work is the core release-rule importance, partial dependence
(survival probability $S(\tau)$ and RMST through the release-rule engine,
plus cumulative hazard via the embedded `$rf` survival forest), and
anomaly scoring on the predictor matrix. For many applied problems,
those three views cover the questions you actually want to answer.

The PBC (primary biliary cirrhosis) dataset from `randomForestSRC` has
418 patients, seven predictors, and a Surv-encoded outcome of days to
event (death or transplant, status ∈ {0, 1, 2}). We use a small
seven-variable subset so the vignette fits quickly. For a full
analysis including time-dependent covariates, @Lee:2021 demonstrates the
boosted nonparametric hazard framework that varPro's survival path draws
on.

```{r pbc-data}
library(survival)
data(pbc, package = "randomForestSRC")
pbc_small <- pbc[, c("days", "status", "age", "albumin", "bili",
                     "edema", "platelet")]
pbc_small <- na.omit(pbc_small)
set.seed(20260527L)
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
v_pbc <- if (is.null(.vp$v_pbc)) {
  varPro::varpro(Surv(days, status) ~ ., data = pbc_small, ntree = 50)
} else {
  .vp$v_pbc
}
```

## Variable importance: `gg_varpro()`

```{r pbc-gg-varpro}
plot(gg_varpro(v_pbc))
```

## Partial dependence: `gg_partial_varpro()` on survival

For survival `gg_partial_varpro()` defaults to **survival probability**
(`scale = "auto"` resolves to `"surv"`): each curve is
$S(\tau \mid x)$, the predicted probability of surviving past a horizon
$\tau$, computed through `partialpro()` on the same release-rule (UVT)
engine as the regression and classification fits — bounded in $[0, 1]$
and read in the model's own time units.

When you do not supply `time`, $\tau$ defaults to the **median follow-up
time** of the fit. Because that horizon is derived from the data it is
always in the model's units and cannot be mis-specified the way a
hand-typed number can — a units mismatch (days vs. years) is the classic
survival partial-plot trap, and a data-driven default sidesteps it. The
resolved $\tau$ is shown in the axis label and reported with a message;
pass `time = tau` to choose another horizon. As on the classification
probability scale, the `causal` contrast is hidden here (use
`scale = "rmst"` or `"mortality"` to see it).

Other survival scales are explicit opt-ins: `scale = "rmst"` gives
restricted mean survival time RMST$(\tau)$ (also on the release-rule
engine, with the same median-follow-up default $\tau$), and
`scale = "mortality"` keeps the unbounded ensemble-mortality score
[@Ishwaran:2007a] — a relative-risk index, *not* a survival probability.

```{r pbc-gg-partial-varpro}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
gg_pd_pbc <- if (is.null(.vp$pd_pbc)) {
  gg_partial_varpro(object = v_pbc)
} else {
  .vp$pd_pbc
}
plot(gg_pd_pbc)
```

## Anomaly scoring: `gg_isopro()` on the X-matrix

Because `isopro()` only sees the predictor matrix, it doesn't care
about the family. The same call from section 3 works here.

```{r pbc-isopro, cache=TRUE}
# Precomputed offline (see precompute_varpro.R); falls back to a live fit.
iso_pbc <- if (is.null(.vp$iso_pbc)) {
  varPro::isopro(data = pbc_small[, c("age", "albumin", "bili", "platelet")],
                 method = "rnd", sampsize = 256, ntree = 50)
} else {
  .vp$iso_pbc
}
plot(gg_isopro(iso_pbc))
```

## Not available for survival: `gg_beta_varpro`, `gg_ivarpro`

`varPro::beta.varpro()` errors on survival fits in the current release
(it only supports `regr` and `class`). `gg_ivarpro()` for survival is
similarly deferred pending design work on the per-rule risk-scaling
story. Both are tracked for v3.1.0.

If you call either on a survival fit you'll get a clear error message
pointing at the deferred work, not a silent miscalculation. The
family-support matrix in §6 records this; the rest of the toolkit
that *does* work on survival (`gg_varpro`, `gg_partial_varpro`,
`gg_isopro` above; `gg_udependent` shown in §3 on the X-matrix).

# Cross-cutting reference

## Family-support matrix

| Wrapper | regr | class | surv | regr+ |
|---|---|---|---|---|
| `gg_partial_varpro` | ✓ | ✓ (prob) | ✓ (S(τ); rmst/mortality/chf) | ✗ (not audited) |
| `gg_varpro` | ✓ | ✓ (`conditional = TRUE`) | ✓ | ✗ (errors) |
| `gg_udependent` | ✓ (uvarpro on X) | ✓ (X) | ✓ (X) | ✓ (X) |
| `gg_isopro` | ✓ (X) | ✓ (X) | ✓ (X) | ✓ (X) |
| `gg_beta_varpro` | ✓ | ✓ | ✗ (upstream stop) | ✗ (deferred) |
| `gg_ivarpro` | ✓ | ✓ | ✗ (deferred) | ✗ (deferred) |

The four wrappers in the lower-right are the v3.1.0 work surface.

## Factor-level ordering

Across `gg_beta_varpro()` and `gg_ivarpro()`, the `variable` column is
stored as a factor whose levels are set by descending aggregate
importance (`mean(|imp|)` summed across classes for classification).
The default plot inherits that ordering, so faceted views show
variables in the same row order across panels. If you re-shape the
frame downstream and want the order preserved, keep `variable` as a
factor rather than coercing to character.

This convention will be propagated to `gg_vimp` and
`plot.gg_varpro(conditional = TRUE)` in a follow-up release.

## Caching the expensive calls

`varPro::beta.varpro()` and `varPro::ivarpro()` are the two heavy
calls. Both wrappers accept a pre-computed fit (`beta_fit`,
`ivarpro_fit`) so you can iterate on selection, observation index, or
cutoff without re-fitting the lasso or the local-importance machinery.
The vignette uses this throughout: every section computes the heavy
fit once in a `cache: true` chunk and re-uses it for every figure.

Provenance carries `precomputed = TRUE` when the cached path was used,
so downstream tooling can tell the two paths apart.

## Provenance shape

`attr(., "provenance")$cutoff` is always a *named numeric vector*
across the toolkit:

- regression: length 1, named `"regr"`
- classification: length K, named with the response factor levels

Downstream code that picks a value should read it as a vector
(`prov$cutoff[[class_name]]` or `prov$cutoff[[1]]`), not as a scalar.
This contract was established in v2.7.3.9012 (PR #98).

# Further reading

The release-rule framework is laid out in @Lu2024varpro. Worked
implementation examples and the full API are at the
[varPro tools reference site](https://www.varprotools.org/articles/getstarted.html).

The ensemble-mortality scale (`scale = "mortality"`), one of the
survival partial-dependence options alongside the default survival
probability, is introduced in @Ishwaran:2007a (the Annals of Applied
Statistics methods paper for random survival forests); the R-package
side is described in @Ishwaran:2008. The boosted nonparametric hazard framework that informs
varPro's survival path (including the treatment of time-dependent
covariates) is in @Lee:2021.

Each wrapper's help page carries a "What this is doing" section that
goes one level deeper than this vignette. The cross-cutting reference
in §6 maps each wrapper to the forest families it supports and notes
which capabilities are deferred to v3.1.0.

# References
