---
title: "Exploring variable importance with varPro"
author: "John Ehrlinger"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    html-math-method: mathjax
    code-fold: false
bibliography: ggRandomForests.bib
vignette: >
  %\VignetteIndexEntry{Exploring variable importance with varPro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  cache      = TRUE,
  cache.path = "varpro_cache/"
)
```

```{r libs}
library(ggRandomForests)
library(varPro)
library(ggplot2)
```

# What varPro is

Random-forest variable importance has, for two decades, mostly meant
*permutation importance*: shuffle a column, measure the drop in OOB
accuracy, repeat. It works, but the score depends on artificial data — the
permuted column — and the answer is unstable when variables are correlated.

varPro [@Lu2024varpro] replaces the shuffling with *release rules*. From a
guided-splitting forest, varPro samples a collection of decision-rule
branches; for each variable in each rule, it compares the response inside
the rule's region with the response after the constraint on that variable
is "released". The size of that change, aggregated over rules and trees,
is the variable's importance. No synthetic columns, no permutation — the
contrast is between two real subsets of the data.

That core machinery feeds several views. **`gg_varpro()`** summarises the
per-tree importance distribution. **`gg_beta_varpro()`** refines those
release-rule contrasts with a per-rule lasso. **`gg_partial_varpro()`**
turns the release machinery into partial-dependence curves.
**`gg_udependent()`** reads cross-variable dependency off a `uvarpro()`
fit. **`gg_isopro()`** scores observations for anomaly using an
isolation-forest variant. **`gg_ivarpro()`** computes per-observation
local importance.

This vignette walks all six wrappers on three worked examples — a
regression problem (Boston housing), a classification problem (iris,
binary and multi-class), and a survival problem (PBC). The closing
section is a one-page reference matrix mapping each wrapper to the
forest families it supports.

# Regression: Boston housing

We start with the Boston housing data: 506 census tracts, 13 numeric
predictors, median home value as the response.

```{r boston-fit}
data("Boston", package = "MASS")
set.seed(20260527L)
v_boston <- varPro::varpro(medv ~ ., data = Boston, ntree = 50)
v_boston
```

`varpro()` builds the importance machinery — release rules, per-tree
scores, the splitweight vector — once. Every subsequent figure in this
section reads off that single fit.

## Per-tree importance with `gg_varpro()`

The first question varPro answers is the same one permutation importance
answers — which variables matter, ranked. `gg_varpro()` extracts the
per-tree importance scores and draws their distribution as a horizontal
boxplot, one row per variable, sorted by median z-score. Variables above
a configurable cutoff (default 0.79) are highlighted.

```{r boston-gg-varpro}
plot(gg_varpro(v_boston))
```

The narrow boxes near the top are the variables varPro is confident
about — every tree agrees they matter. Wide boxes that straddle the
cutoff line are the ones to look at twice; the forest disagrees with
itself.

## Partial dependence with `gg_partial_varpro()`

The ranking view tells you *which* variables matter. Partial dependence
asks the next question: *how* does the response change with a variable,
holding the others fixed? `gg_partial_varpro()` wraps
`varPro::partialpro()` and returns a tidy frame of parametric,
non-parametric, and causal partial-dependence curves.

```{r boston-gg-partial-varpro}
gg_pd <- gg_partial_varpro(object = v_boston)
plot(gg_pd)
```

Each panel is a single predictor. The three curves correspond to the
three estimators varPro carries — read them as a sensitivity analysis.
Tight agreement is reassuring; divergence is information.
